Possible Inconsistency in Maximum Likelihood Calculation of Ancestral States

### 2017-03-30 joint_inf.tex
a lot of notation updates

% Commands
\newcommand{\alphabet}{\mathcal{A}}
\newcommand{\fullAlignment}{\mathbf{Y}}
\newcommand{\alignmentColumn}{\mathbf{y}}
\newcommand{\sitePattern}{s}
\newcommand{\siteSplit}{S'}
\newcommand{\sitePatternSet}{S}
\newcommand{\fullSitePatterns}{\mathbf{s}}
\newcommand{\sitePatternSpace}{\mathcal{S}}
\newcommand{\fullAncestralStates}{\mathbf{H}}
\newcommand{\ancestralStateSpace}{\mathcal{H}}
\newcommand{\ancestralStateColumn}{\boldsymbol\eta}
\newcommand{\ancestralStateCategory}{Z}
\newcommand{\ancestralSplit}{H'}
\newcommand{\ancestralStateSet}{H}
\newcommand{\fullAncestralStateCategories}{\mathbb{Z}}
\newcommand{\ancestralStateCategorySpace}{\mathcal{Z}}

\newcommand{\nCols}{n}
\newcommand{\nSiteRows}{m}
\newcommand{\nAncestralStateRows}{p}
\newcommand{\nSiteSplits}{q}
\newcommand{\nAncestralSplits}{r}
\newcommand{\nMostLikelyAncestralSplits}{\tilde{r}}

@@

We will be considering phylogenetic likelihood on unrooted phylogenetic trees.
We assume the character alphabet $\alphabet=\{0,1\}$ and a uniform stationary distribution.
Let $\nSiteRows$ be the number of tips of the tree, and $\nAncestralStateRows = \nSiteRows-2$ the number of internal nodes.
%EM Commented out below because it's not right-- fullAlignment is not the set of assignments of characters to tips of the tree, it's a single such.
%Let $\fullAlignment$ be the set of assignments of characters to tips of our tree, and $\fullAncestralStates$ the set of assignments of characters to the internal nodes of the tree.
Assume we observe $\nCols$ samples of character data (i.e.\ an alignment with $\nCols$ columns) $\fullAlignment=[\alignmentColumn_1,\ldots,\alignmentColumn_\nCols]\in\alphabet^{\nSiteRows\times\nCols}$.
There are also corresponding unknown ancestral states $\fullAncestralStates=[\ancestralStateColumn_1,\ldots,\ancestralStateColumn_\nCols]\in\alphabet^{\nAncestralStateRows\times\nCols}$.

@@

For a topology $\tau$ and branch lengths $t$ we start with the fully-observed likelihood (making the usual independence assumption between sites):
\begin{equation}
\label{eq:full_likelihood}
L_\nCols(\fullAlignment;\fullAncestralStates, \tau, t) = \prod_{i=1}^{\nCols} \ P(\alignmentColumn_i, \ancestralStateColumn_i | \tau, t).
\end{equation}
% In particular, we are interested in
% $$
% (\hat{\boldsymbol\xi}, \hat{\tau}, \hat{t}) = \arg\max_{\boldsymbol\xi, \tau, t} \ L_n(\mathbf{y};\boldsymbol\xi, \tau, t),
% $$
% which we call the maximum likelihood values of the parameters $(\boldsymbol\xi, \tau, t)$.
% Since the number of elements in $\boldsymbol\xi$ grows with that of the observed data $\mathbf{y}$,
The typical approach to estimate the tree $\tau$ and branch lengths $t$ involves marginalizing across ancestral states
\begin{equation}
\label{eq:marginal_likelihood}
\tilde{L}_\nCols(\fullAlignment; \tau, t) = \sum_{\fullAncestralStates} \ L_\nCols(\fullAlignment;\fullAncestralStates, \tau, t)
\end{equation}
and maximizing over the topology and branch lengths to obtain
$$
(\hat{\tau}, \hat{t}) = \arg\max_{\tau, t} \  \tilde{L}_\nCols(\fullAlignment; \tau, t).
$$
% The values $\hat{\boldsymbol\xi}$ are then calculated conditional on these estimates.
% More likely in practice we fix a topology $\tau$ and use this marginalization approach to compute $(\hat{\boldsymbol\xi}, \hat{t})$.

@@

Since we have a finite character alphabet, for a given column $i$ there are a finite number of possibilities of assignments of characters to tips $\alignmentColumn_i$ or internal nodes $\ancestralStateColumn_i$.
Following Section 8.6 of \cite{Semple2003-em}, $\tau$ induces an ordered set of tip labels $\{1,\ldots,\nSiteRows\}$ across sites.
%EM What are complementary characters?
For a binary character model, a realization of $\alignmentColumn_i$ can be described as a subset of complementary characters $\siteSplit\subseteq\sitePatternSet\triangleq\{1,\ldots,\nSiteRows-1\}$.
%EM Why nSiteRows-1 in the def of siteSplit above, but nAncestralStateRows below in the def of ancestralSplit?
%EM There are two ways to do this, if you are collapsing such that 001 and 110 are the same or not, and we should be clear which one you are doing.
The topology $\tau$ induces an ordered set $\{1,\ldots,\nAncestralStateRows\}$ for $\ancestralStateColumn_i$ with a similar subset of complementary characters $\ancestralSplit\subseteq\ancestralStateSet\triangleq\{1,\ldots,\nAncestralStateRows\}$ describing a realization of $\ancestralStateColumn_i$.
Define the total number of unique site splits as $\nSiteSplits=|\mathcal{P}(\sitePatternSet)|$ and unique ancestral state splits as $\nAncestralSplits=|\mathcal{P}(\ancestralStateSet)|$ where $\mathcal{P}$ denotes the powerset.

%das: Maybe writing out the example would help?
%EM: Yes, I think so. e.g. the need for partitions \ancestralStateCategory may not be obvious right away.

Consider the probability $P(\alignmentColumn_i, \ancestralStateColumn_i | \tau, t)$ from \eqref{eq:full_likelihood} of observing a particular column $i$.
%EM Indexing \siteSplit_j$ not defined.
%EM But, more importantly, as we discussed Z_j is defined only with respect to a single tree and branch lengths. This need to be made abundantly clear. Personally, I think that using the same index j for them is confusing.
Given $(\tau, t)$ and maximizing over $\ancestralStateColumn_i$, if $\alignmentColumn_i=\siteSplit_j$ then there exists a partition $\ancestralStateCategory_j$ of $\mathcal{P}(\ancestralStateSet)$ of most likely ancestral states where
\begin{align*}
\max_{\ancestralSplit_k\in\mathcal{P}(\ancestralStateSet)} \ P(\alignmentColumn_i=\siteSplit_j, \ancestralStateColumn_i=\ancestralSplit_k | \tau, t) &= P(\alignmentColumn_i=\siteSplit_j, \ancestralStateColumn_i\in\ancestralStateCategory_j | \tau, t) \\
     &= P(\alignmentColumn=\siteSplit_j, \ancestralStateColumn\in\ancestralStateCategory_j | \tau, t).
\end{align*}
This equivalence is helpful in writing the full likelihood of the joint inference problem in terms of site splits; for most likely ancestral state partitions $\fullAncestralStateCategories=\{\ancestralStateCategory_1,\ldots,\ancestralStateCategory_\nSiteSplits\}$ we reformulate the likelihood equation
\begin{align}
L_\nCols'(\fullAlignment;\tau, t) &= \max_{\fullAncestralStates} \ L_\nCols(\fullAlignment;\fullAncestralStates, \tau, t) \\
                             &= \max_{\fullAncestralStates} \ \prod_{i=1}^{\nCols} \ P(\alignmentColumn_i, \ancestralStateColumn_i | \tau, t) \\
                             &= \prod_{i=1}^{\nCols} \ \max_{\ancestralSplit_k\in\mathcal{P}(\ancestralStateSet)} \ P(\alignmentColumn_i=\siteSplit_j, \ancestralStateColumn_i=\ancestralSplit_k | \tau, t) \\
%EM Can you think a little about z_j?
%EM Even though it's a category now, it's still a function of tau and t, right?
                             &= \prod_{i=1}^{\nCols} \ P(\alignmentColumn=\siteSplit_j, \ancestralStateColumn\in\ancestralStateCategory_j | \tau, t) \\
                             &= \prod_{j=1}^{\nSiteSplits} \ P(\alignmentColumn=\siteSplit_j, \ancestralStateColumn\in\ancestralStateCategory_j | \tau, t)^{\nCols_j(\fullAlignment)} \label{eq:site_pattern_likelihood}
\end{align}
where $\nCols_j(\fullAlignment)$ is the number of columns in $\fullAlignment$ where the site split $\siteSplit_j$ appears.

@@

We are interested in the properties of the objective function being maximized in \eqref{eq:site_pattern_likelihood} as data becomes plentiful.
Define \eqref{eq:site_pattern_likelihood} as
$$
L_\nCols''(\fullAlignment;\fullAncestralStateCategories,\tau,t) = \prod_{j=1}^{\nSiteSplits} \ P(\alignmentColumn=\siteSplit_j, \ancestralStateColumn\in\ancestralStateCategory_j | \tau, t)^{\nCols_j(\fullAlignment)} .
$$
Assume our $\nCols$ observations were generated from a model with parameters $(\tau^*, t^*)$.
We have
\begin{align}
    \frac{1}{\nCols} \log L_\nCols''(\fullAlignment;\fullAncestralStateCategories,\tau,t)
        &= \sum_{j}^\nSiteSplits \frac{\nCols_j(\fullAlignment)}{\nCols}\cdot  \log P(\alignmentColumn=\siteSplit_j, \ancestralStateColumn\in\ancestralStateCategory_j | \tau, t) \\
        &= \sum_{j}^\nSiteSplits \frac{\nCols_j(\fullAlignment)}{\nCols}\cdot [\log P(\alignmentColumn=\siteSplit_j | \tau, t) +
            \log P(\ancestralStateColumn\in\ancestralStateCategory_j | \alignmentColumn=\siteSplit_j , \tau, t)]
\end{align}
so that, in the limit of infinite data,
\begin{equation}
\frac{1}{\nCols} \log L_\nCols''(\fullAlignment;,\tau,t) \rightarrow \sum_{j}^\nSiteSplits P(\alignmentColumn=\siteSplit_j | \tau^*, t^*) \cdot [\log P(\alignmentColumn=\siteSplit_j | \tau, t) + \log P(\ancestralStateColumn\in\ancestralStateCategory_j | \alignmentColumn=\siteSplit_j , \tau, t)]. \label{eq:site_pattern_profile_likelihood_mean}
\end{equation}
Define the the information quantity
$$
h_{\tau^*,t^*}(\tau,t) = \sum_{j}^\nSiteSplits P(\alignmentColumn=\siteSplit_j | \tau^*, t^*) \log P(\siteSplit_j | \tau, t),
$$
and the partial log likelihood
$$
\ell^p_{\tau^*,t^*}(\fullAncestralStateCategories,\tau,t) = \sum_{j}^\nSiteSplits P(\alignmentColumn=\siteSplit_j | \tau^*, t^*) \log P(\ancestralStateColumn\in\ancestralStateCategory_j | \siteSplit_j, \tau, t)
$$
so that \eqref{eq:site_pattern_profile_likelihood_mean} is written as
\begin{equation}
    \label{eq:log_likelihood_simplified}
\ell_{\tau^*,t^*}(\fullAncestralStateCategories,\tau,t) = h_{\tau^*,t^*}(\tau,t) + \ell^p_{\tau^*,t^*}(\fullAncestralStateCategories,\tau,t).
\end{equation}
Recall $(\tau, t)$ are estimands and $(\tau^*, t^*)$ are fixed, generating parameters.
For data generated from $(\tau^*, t^*)$, if
\begin{equation}
\label{eq:inconsistency_inequality}
\ell_{\tau^*,t^*}(\hat{\fullAncestralStateCategories}_1,\tau',\hat{t}_1) > \ell_{\tau^*,t^*}(\hat{\fullAncestralStateCategories}_2,\tau^*,\hat{t}_2)
\end{equation}
for $\tau'\neq\tau^*$ with $\{\hat{t}_1,\hat{\fullAncestralStateCategories}_1\}$ and $\{\hat{t}_2,\hat{\fullAncestralStateCategories}_2\}$ estimated by maximizing \eqref{eq:profile_likelihood}, we have an inconsistency.

@@

Our approach will be to seek bounds for the terms inside the maximization in \eqref{eq:log_likelihood_simplified} as
\begin{equation}
\label{eq:partial-likelihood}
\ell_{\tau^*,t^*}(\fullAncestralStateCategories,\tau,t) = h_{\tau^*,t^*}(\tau,t) + \ell^p_{\tau^*,t^*}(\fullAncestralStateCategories,\tau,t).
\end{equation}
We obtain an upper bound for the joint maximum of \eqref{eq:partial-likelihood}
$$
\max_{t,\fullAncestralStateCategories} \ \ell_{\tau^*,t^*}(\fullAncestralStateCategories,\tau,t) \le
    h_{\tau^*,t^*}(\tau^*,t^*)
    + \ell^p_{\tau^*,t^*}(\hat{\fullAncestralStateCategories},\tau,\hat{t})
$$
using Gibbs's inequality
$$
h_{\tau^*,t^*}(\tau,t) \le h_{\tau^*,t^*}(\tau^*,t^*)
$$
and
$$
\{\hat{t},\hat{\fullAncestralStateCategories}\} = \arg\max_{t,\fullAncestralStateCategories} \ \ell^p_{\tau^*,t^*}(\fullAncestralStateCategories,\tau,t).
$$
Similarly, we have a lower bound as
$$
\max_{t,\fullAncestralStateCategories} \ \ell_{\tau^*,t^*}(\fullAncestralStateCategories,\tau,t) \ge
    h_{\tau^*,t^*}(\tau,\hat{t})
    + \ell^p_{\tau^*,t^*}(\hat{\fullAncestralStateCategories},\tau,\hat{t}).
$$

@@

    $\siteSplit_j$    & $\ancestralStateCategory_j$ & $P(\ancestralStateColumn\in\ancestralStateCategory_j|\alignmentColumn=\siteSplit_j,\tau,t)$\\

@@

\caption{Likelihood calculations for all site patterns and ancestral state categories.

>>>>

% Commands
\newcommand{\alphabet}{\mathcal{A}}
\newcommand{\fullAlignment}{\mathbf{Y}}
\newcommand{\alignmentColumn}{\mathbf{y}}
\newcommand{\alignmentColumnRV}{Y}
\newcommand{\siteSplit}{\tilde{s}}
\newcommand{\siteSplitSet}{\mathcal{S}}
\newcommand{\fullAncestralStates}{\mathbf{H}}
\newcommand{\ancestralStateColumn}{\mathbf{h}}
\newcommand{\ancestralStateColumnRV}{H}
\newcommand{\ancestralSplit}{\tilde{h}}
\newcommand{\ancestralSplitSet}{\mathcal{H}}
\newcommand{\ancestralSplitPartition}{\eta}
\newcommand{\fullAncestralSplitPartitions}{\boldsymbol\eta}

\newcommand{\patternToSplit}{\psi}
\newcommand{\ancestralToSplit}{\xi}

\newcommand{\siteSplitRV}{\Psi}
\newcommand{\ancestralSplitRV}{\Xi}

\newcommand{\nCols}{n}
\newcommand{\nSiteRows}{m}
\newcommand{\nAncestralStateRows}{p}
\newcommand{\nSiteSplits}{q}
\newcommand{\nAncestralSplits}{r}

\newcommand{\shannonDivergence}{D}

\allowdisplaybreaks

@@

We will be considering phylogenetic likelihood on unrooted phylogenetic trees.
We assume the character alphabet $\alphabet=\{0,1\}$ and a uniform stationary distribution.
Let $\nSiteRows$ be the number of tips of the tree, and $\nAncestralStateRows = \nSiteRows-2$ the number of internal nodes.
%EM Commented out below because it's not right-- fullAlignment is not the set of assignments of characters to tips of the tree, it's a single such.
%Let $\fullAlignment$ be the set of assignments of characters to tips of our tree, and $\fullAncestralStates$ the set of assignments of characters to the internal nodes of the tree.
Assume we observe $\nCols$ samples of character data (i.e.\ an alignment with $\nCols$ columns) $\fullAlignment=[\alignmentColumn_1,\ldots,\alignmentColumn_\nCols]\in\alphabet^{\nSiteRows\times\nCols}$ distributed as the random variable $\alignmentColumnRV$.
There are also corresponding unknown ancestral states $\fullAncestralStates=[\ancestralStateColumn_1,\ldots,\ancestralStateColumn_\nCols]\in\alphabet^{\nAncestralStateRows\times\nCols}$ distributed as $\ancestralStateColumnRV$.

@@

For a topology $\tau$ and branch lengths $t$ we start with the fully-observed likelihood (making the usual independence assumption between sites):
\begin{equation}
\label{eq:full_likelihood}
L_\nCols(\fullAlignment;\fullAncestralStates, \tau, t) = \prod_{i=1}^{\nCols} \ P(\alignmentColumnRV=\alignmentColumn_i, \ancestralStateColumnRV=\ancestralStateColumn_i | \tau, t).
\end{equation}
% In particular, we are interested in
% $$
% (\hat{\boldsymbol\xi}, \hat{\tau}, \hat{t}) = \arg\max_{\boldsymbol\xi, \tau, t} \ L_n(\mathbf{y};\boldsymbol\xi, \tau, t),
% $$
% which we call the maximum likelihood values of the parameters $(\boldsymbol\xi, \tau, t)$.
% Since the number of elements in $\boldsymbol\xi$ grows with that of the observed data $\mathbf{y}$,
The typical approach to estimate the tree $\tau$ and branch lengths $t$ involves marginalizing across all possible ancestral states
\begin{equation}
\label{eq:marginal_likelihood}
\tilde{L}_\nCols(\fullAlignment; \tau, t) = \sum_{\fullAncestralStates\in\alphabet^{\nAncestralStateRows\times\nCols}} \ L_\nCols(\fullAlignment;\fullAncestralStates, \tau, t)
\end{equation}
and maximizing over the topology and branch lengths to obtain
$$
(\hat{\tau}, \hat{t}) = \arg\max_{\tau, t} \  \tilde{L}_\nCols(\fullAlignment; \tau, t).
$$
% The values $\hat{\boldsymbol\xi}$ are then calculated conditional on these estimates.
% More likely in practice we fix a topology $\tau$ and use this marginalization approach to compute $(\hat{\boldsymbol\xi}, \hat{t})$.

@@

Since we have a finite character alphabet, for a given column $i$ there are a finite number of possibilities of assignments of characters to tips $\alignmentColumn_i$ or internal nodes $\ancestralStateColumn_i$.
Following Section 8.6 of \cite{Semple2003-em}, $\tau$ induces an ordered set of tip labels $\{1,\ldots,\nSiteRows\}$ across sites.
For a binary character model, a realization of $\alignmentColumn_i$ can be described as a subset of indices $\siteSplit\subseteq\siteSplitSet\triangleq\{1,\ldots,\nSiteRows-1\}$ with equivalent characters, commonly called a ``site split.''
In the binary character setting, the probability of observing a particular site split is equivalent to the probability of its complement, resulting in one fewer index in $\siteSplitSet$ than the number of tips.
Define the total number of unique site splits as $\nSiteSplits=|\mathcal{P}(\siteSplitSet)|$ where $\mathcal{P}$ denotes the powerset.
The topology $\tau$ also induces an ordered set of internal nodes $\{1,\ldots,\nAncestralStateRows\}$ for $\ancestralStateColumn_i$ with a similar subset of characters $\ancestralSplit\subseteq\ancestralSplitSet\triangleq\{1,\ldots,\nAncestralStateRows\}$ describing a realization of $\ancestralStateColumn_i$.
The probability of observing an ancestral state split, conditional on a site split, will not be invariant to taking its complement, and thus the need for enumerating the entire set of internal nodes.
The number of unique ancestral state splits is $\nAncestralSplits=|\mathcal{P}(\ancestralSplitSet)|$.

%EM Why nSiteRows-1 in the def of siteSplit above, but nAncestralStateRows below in the def of ancestralSplit?
%EM There are two ways to do this, if you are collapsing such that 001 and 110 are the same or not, and we should be clear which one you are doing.

With this split formulation, define the mappings from patterns to splits as
$$
\patternToSplit:\alphabet^\nSiteRows\rightarrow\mathcal{P}(\siteSplitSet), \ \ancestralToSplit:\alphabet^\nAncestralStateRows\rightarrow\mathcal{P}(\ancestralSplitSet)
$$
so that, for
$$
\patternToSplit(\alignmentColumn_i)=\siteSplit_j, \ \ancestralToSplit(\ancestralStateColumn_i)=\ancestralSplit_k
$$
and
$$
\patternToSplit(\alignmentColumnRV)=\siteSplitRV, \ \ancestralToSplit(\ancestralStateColumnRV)=\ancestralSplitRV,
$$
we write the $i$th factor of \eqref{eq:full_likelihood} as
$$
P(\alignmentColumnRV=\alignmentColumn_i, \ancestralStateColumnRV=\ancestralStateColumn_i | \tau, t)=P(\siteSplitRV=\siteSplit_j, \ancestralSplitRV=\ancestralSplit_k | \tau, t).
$$
Taking the maximum of the probability of observing a particular column $i$ yields
\begin{align*}
\max_{\ancestralSplit_k\in\mathcal{P}(\ancestralSplitSet)} \ P(\siteSplitRV=\siteSplit_j, \ancestralSplitRV=\ancestralSplit_k | \tau, t) &=
\max_{\ancestralSplit_k\in\mathcal{P}(\ancestralSplitSet)} \ P(\siteSplitRV=\siteSplit_j| \tau, t)\cdot P(\ancestralSplitRV=\ancestralSplit_k |\siteSplitRV=\siteSplit_j, \tau, t) \\
&= P(\siteSplitRV=\siteSplit_j| \tau, t)\cdot\max_{\ancestralSplit_k\in\mathcal{P}(\ancestralSplitSet)} \ P(\ancestralSplitRV=\ancestralSplit_k |\siteSplitRV=\siteSplit_j, \tau, t).
\end{align*}
%EM Indexing \siteSplit_j$ not defined.
%EM But, more importantly, as we discussed Z_j is defined only with respect to a single tree and branch lengths. This need to be made abundantly clear. Personally, I think that using the same index j for them is confusing.
Given $(\tau, t)$ and $\siteSplitRV=\siteSplit_j$, there exists a partition $\ancestralSplitPartition(\siteSplit_j, \tau, t)=\ancestralSplitPartition_j(\tau, t)$ of $\mathcal{P}(\ancestralSplitSet)$ of most likely ancestral states where
\begin{align*}
\max_{\ancestralSplit_k\in\mathcal{P}(\ancestralSplitSet)} \ P(\ancestralSplitRV=\ancestralSplit_k | \siteSplitRV=\siteSplit_j, \tau, t) &= P(\ancestralSplitRV\in\ancestralSplitPartition_j(\tau, t) | \siteSplitRV=\siteSplit_j, \tau, t) \\
     &= P(\ancestralSplitRV\in\ancestralSplitPartition_j(\tau, t) | \siteSplitRV=\siteSplit_j, \tau, t).
\end{align*}

For most likely ancestral state partitions $\fullAncestralSplitPartitions(\tau, t)=\{\ancestralSplitPartition_1(\tau, t),\ldots,\ancestralSplitPartition_\nSiteSplits(\tau, t)\}$ we reformulate the likelihood equation
\begin{align}
L_\nCols'(\fullAlignment;\tau, t) &= \max_{\fullAncestralStates} \ L_\nCols(\fullAlignment;\fullAncestralStates, \tau, t) \\
                             &= \max_{\fullAncestralStates} \ \prod_{i=1}^{\nCols} \ P(\alignmentColumnRV=\alignmentColumn_i, \ancestralStateColumnRV=\ancestralStateColumn_i | \tau, t) \\
                             &= \prod_{i=1}^{\nCols} \ \max_{\ancestralSplit_k\in\mathcal{P}(\ancestralSplitSet)} \ P(\siteSplitRV=\siteSplit_j, \ancestralSplitRV=\ancestralSplit_k | \tau, t) \\
%EM Can you think a little about z_j?
%EM Even though it's a category now, it's still a function of tau and t, right?
                             &= \prod_{i=1}^{\nCols} \ P(\siteSplitRV=\siteSplit_j| \tau, t)\cdot\max_{\ancestralSplit_k\in\mathcal{P}(\ancestralSplitSet)} \ P(\ancestralSplitRV=\ancestralSplit_k |\siteSplitRV=\siteSplit_j, \tau, t) \\
                             &= \prod_{i=1}^{\nCols} \ P(\siteSplitRV=\siteSplit_j | \tau, t)\cdot P(\ancestralSplitRV\in\ancestralSplitPartition_j(\tau, t) | \siteSplitRV=\siteSplit_j, \tau, t) \\
                             &= \prod_{j=1}^{\nSiteSplits} \ \left[P(\siteSplitRV=\siteSplit_j | \tau, t)\cdot P(\ancestralSplitRV\in\ancestralSplitPartition_j(\tau, t) | \siteSplitRV=\siteSplit_j, \tau, t)\right] ^{\nCols_j(\fullAlignment)} \label{eq:site_pattern_likelihood}
\end{align}
where $\nCols_j(\fullAlignment)$ is the number of columns in $\fullAlignment$ where the site split $\siteSplit_j$ appears.

\paragraph{Example}

Consider a fixed, binary four-tip tree $\tau$.
The space of all possible character assignments, in terms of site splits, is
\begin{align*}
\mathcal{P}(\{1,2,3,4\}) &= \{\{\emptyset\}, \{1,2,3,4\}, \{1\}, \{2,3,4\}, \{2\}, \{1,3,4\}, \{3\}, \{1,2,4\}, \\
                         &\qquad \{1,2\}, \{3,4\}, \{1,3\}, \{2,4\}, \{2,3\}, \{1,4\}, \{1,2,3\}, \{1,4\}\}.
\end{align*}
where each set indicates the tips at which a character is assigned the character $1$.
For example, $\{\emptyset\}$ is the labeling $0000$ and $\{1,3,4\}$ is the labeling $1011$.
Symmetry allows us to group adjacent pairs in $\mathcal{P}(\{1,2,3,4\})$ into equiprobable splits, letting $\siteSplitSet=\{1,2,3\}$.
The unique site splits, collapsing complements, are
$$
\mathcal{P}(\siteSplitSet) = \{\{\emptyset\}, \{1\}, \{2\}, \{3\}, \{1,2\}, \{1,3\}, \{2,3\}, \{1,2,3\}\}.
$$
Had we not identified character complements, we would need to consider the additional splits
$$
\mathcal{P}(\{1,2,3,4\}) \setminus \mathcal{P}(\siteSplitSet) = \{\{1,2,3,4\}, \{2,3,4\}, \{1,3,4\}, \{1,2,4\}, \{3,4\}, \{2,4\}, \{1,4\}, \{4\}\},
$$
but the symmetry of the binary character model allows us to focus just on the elements of $\mathcal{P}(\siteSplitSet)$.
This tree has two internal nodes with $\ancestralSplitSet=\{1,2\}$ and unique ancestral state splits
$$
\mathcal{P}(\ancestralSplitSet) = \{\{\emptyset\}, \{1\}, \{2\}, \{1,2\}\}.
$$
We cannot similarly identify complements in this case since, for $\siteSplit\in\mathcal{P}(\siteSplitSet)$,
$$
P(\ancestralSplitRV=\{\emptyset\}|\siteSplitRV=\siteSplit, \tau, t)\neq P(\ancestralSplitRV=\{1,2\}|\siteSplitRV=\siteSplit, \tau, t)
$$
in general.

For each site split $\siteSplit\in\mathcal{P}(\siteSplitSet)$, we maximize the likelihood over all $\ancestralSplit\in\mathcal{P}(\ancestralSplitSet)$.
A maximum will occur at one of possibly several ancestral splits in $\mathcal{P}(\ancestralSplitSet)$; define $\ancestralSplitPartition(\siteSplit, \tau, t)$ as the partition of $\mathcal{P}(\ancestralSplitSet)$ into those most likely ancestral splits for that particular site split, topology and set of branch lengths.
Assume all branch lengths correspond to a probability $p$ ($< 1/2$) of changing character along that branch.
The probabilities of observing ancestral splits for site split $\{\emptyset\}$ are 
$$
P(\ancestralSplitRV=\{\emptyset\}|\siteSplitRV=\{\emptyset\}, \tau, t) = 
(1-p)^5,
$$
$$
P(\ancestralSplitRV=\{1\}|\siteSplitRV=\{\emptyset\}, \tau, t) =
P(\ancestralSplitRV=\{2\}|\siteSplitRV=\{\emptyset\}, \tau, t) = 
p^2(1-p)^3,
$$
$$
P(\ancestralSplitRV=\{1,2\}|\siteSplitRV=\{\emptyset\}, \tau, t) = 
p^4(1-p).
$$
Here, the partition into most likely ancestral states will proceed as
$$
P(\ancestralSplitRV\in\ancestralSplitPartition_j(\tau, p)|\siteSplitRV=\{\emptyset\}, \tau, t) = 
(1-p)^5
$$
where the partition is the single element $\ancestralSplitPartition_j(\tau, p)=\{\emptyset\}$.
For the site split $\{1,3\}$ we have
$$
P(\ancestralSplitRV=\{\emptyset\}|\siteSplitRV=\{1,3\}, \tau, t) = 
P(\ancestralSplitRV=\{1,2\}|\siteSplitRV=\{1,3\}, \tau, t) =
p^2(1-p)^3,
$$
$$
P(\ancestralSplitRV=\{1\}|\siteSplitRV=\{1,3\}, \tau, t) = 
P(\ancestralSplitRV=\{2\}|\siteSplitRV=\{1,3\}, \tau, t) =
p^3(1-p)^2,
$$
where
$$
P(\ancestralSplitRV\in\ancestralSplitPartition_j(\tau, p)|\siteSplitRV=\{1,3\}, \tau, t) = 
p^2(1-p)^3
$$
with the most likely partition $\ancestralSplitPartition_j(\tau, p)=\{\{\emptyset\},\{1,2\}\}.$

@@

We are interested in the properties of the objective function being maximized in \eqref{eq:site_pattern_likelihood} as data becomes plentiful.
Define \eqref{eq:site_pattern_likelihood} as
$$
L_\nCols''(\fullAlignment;\fullAncestralSplitPartitions(\tau, t),\tau,t) = \prod_{j=1}^{\nSiteSplits} \ \left[P(\siteSplitRV=\siteSplit_j | \tau, t) \cdot P(\ancestralSplitRV\in\ancestralSplitPartition_j(\tau, t) | \siteSplitRV=\siteSplit_j, \tau, t)\right] ^{\nCols_j(\fullAlignment)} \label{eq:site_pattern_likelihood}
$$
Assume our $\nCols$ observations were generated from a model with parameters $(\tau^*, t^*)$.
We have
\begin{align}
    \frac{1}{\nCols} \log L_\nCols''(\fullAlignment;\fullAncestralSplitPartitions(\tau,t),\tau,t)
        &= \sum_{j=1}^\nSiteSplits \frac{\nCols_j(\fullAlignment)}{\nCols}\cdot  \log P(\siteSplitRV=\siteSplit_j, \ancestralSplitRV\in\ancestralSplitPartition_j(\tau, t) | \tau, t) \\
        &= \sum_{j=1}^\nSiteSplits \frac{\nCols_j(\fullAlignment)}{\nCols}\cdot [\log P(\siteSplitRV=\siteSplit_j | \tau, t) +
            \log P(\ancestralSplitRV\in\ancestralSplitPartition_j(\tau, t) | \siteSplitRV=\siteSplit_j , \tau, t)]
\end{align}
so that, in the limit of infinite data,
\begin{equation}
\frac{1}{\nCols} \log L_\nCols''(\fullAlignment;,\tau,t) \rightarrow \sum_{j=1}^\nSiteSplits P(\siteSplitRV=\siteSplit_j | \tau^*, t^*) \cdot [\log P(\siteSplitRV=\siteSplit_j | \tau, t) + \log P(\ancestralSplitRV\in\ancestralSplitPartition_j(\tau, t) | \siteSplitRV=\siteSplit_j , \tau, t)]. \label{eq:site_pattern_profile_likelihood_mean}
\end{equation}
Define the the divergence quantity
$$
\shannonDivergence_{\tau^*,t^*}(\tau,t) = \sum_{j=1}^\nSiteSplits P(\siteSplitRV=\siteSplit_j | \tau^*, t^*)\cdot\log P(\siteSplitRV=\siteSplit_j | \tau, t),
$$
and the partial log likelihood
$$
\ell^p_{\tau^*,t^*}(\fullAncestralSplitPartitions(\tau,t),\tau,t) = \sum_{j=1}^\nSiteSplits P(\siteSplitRV=\siteSplit_j | \tau^*, t^*)\cdot\log P(\ancestralSplitRV\in\ancestralSplitPartition_j(\tau, t) | \siteSplit_j, \tau, t)
$$
so that \eqref{eq:site_pattern_profile_likelihood_mean} is written as
\begin{equation}
    \label{eq:log_likelihood_simplified}
    \ell_{\tau^*,t^*}(\fullAncestralSplitPartitions(\tau,t),\tau,t) = \shannonDivergence_{\tau^*,t^*}(\tau,t) + \ell^p_{\tau^*,t^*}(\fullAncestralSplitPartitions(\tau,t),\tau,t).
\end{equation}
Recall $(\tau, t)$ are estimands and $(\tau^*, t^*)$ are fixed, generating parameters.
For data generated from $(\tau^*, t^*)$, if
\begin{equation}
\label{eq:inconsistency_inequality}
\ell_{\tau^*,t^*}(\widehat{\fullAncestralSplitPartitions(\tau',t_1)},\tau',\hat{t}_1) > \ell_{\tau^*,t^*}(\widehat{\fullAncestralSplitPartitions(\tau^*,t_2)},\tau^*,\hat{t}_2)
\end{equation}
for $\tau'\neq\tau^*$ with $\{\hat{t}_1,\widehat{\fullAncestralSplitPartitions(\tau',t_1)}\}$ and $\{\hat{t}_2,\widehat{\fullAncestralSplitPartitions(\tau^*,t_2)}\}$ estimated by maximizing \eqref{eq:profile_likelihood}, we have an inconsistency.

@@

Our approach will be to seek bounds for the terms inside the maximization in \eqref{eq:log_likelihood_simplified} as
\begin{equation}
\label{eq:partial-likelihood}
\ell_{\tau^*,t^*}(\fullAncestralSplitPartitions(\tau,t),\tau,t) = \shannonDivergence_{\tau^*,t^*}(\tau,t) + \ell^p_{\tau^*,t^*}(\fullAncestralSplitPartitions(\tau,t),\tau,t).
\end{equation}
We obtain an upper bound for the joint maximum of \eqref{eq:partial-likelihood}
$$
\max_{t,\fullAncestralSplitPartitions(\tau,t)} \ \ell_{\tau^*,t^*}(\fullAncestralSplitPartitions(\tau,t),\tau,t) \le
    \shannonDivergence_{\tau^*,t^*}(\tau^*,t^*)
    + \ell^p_{\tau^*,t^*}(\widehat{\fullAncestralSplitPartitions(\tau,t)},\tau,\hat{t})
$$
using Gibbs's inequality
$$
\shannonDivergence_{\tau^*,t^*}(\tau,t) \le \shannonDivergence_{\tau^*,t^*}(\tau^*,t^*)
$$
and
$$
\{\hat{t},\widehat{\fullAncestralSplitPartitions(\tau,t)}\} = \arg\max_{t,\fullAncestralSplitPartitions(\tau,t)} \ \ell^p_{\tau^*,t^*}(\fullAncestralSplitPartitions(\tau,t),\tau,t).
$$
Similarly, we have a lower bound as
$$
\max_{t,\fullAncestralSplitPartitions(\tau,t)} \ \ell_{\tau^*,t^*}(\fullAncestralSplitPartitions(\tau,t),\tau,t) \ge
    \shannonDivergence_{\tau^*,t^*}(\tau,\hat{t})
    + \ell^p_{\tau^*,t^*}(\widehat{\fullAncestralSplitPartitions(\tau,t)},\tau,\hat{t}).
$$

@@

    $\siteSplit_j$    & $\ancestralSplitPartition_j(\tau, t)$ & $P(\ancestralSplitRV\in\ancestralSplitPartition_j(\tau, t)|\siteSplitRV=\siteSplit_j,\tau,t)$\\

@@

\caption{Likelihood calculations for all site patterns and ancestral state partitions.
